Partial and total correctness \\ as greatest and least fixed points

### 2015-01-21 paper.tex
+

In the context of Hoare Logic~\cite{hoare69}, a (possibly
non-deterministic, possibly non-terminating) program satisfies a
\emph{total correctness} specification when each of its traces from a
state that satisfies the given precondition reaches a terminal state
satisfying the given postcondition within a finite number of execution
steps. That program satisfies a \emph{partial correctness}
specification when each of its traces either meets the requirement
above or is infinite.

@@

The descriptions above have been deliberately crafted to emphasise a
connection between total/partial correctness and least/greatest fixed
points. This paper seeks to state this connection precisely, and
investigate conditions under which it holds.

Section~\ref{sec:result} provides characterisations of partial and
total correctness, that differ only in that the former takes a
function's greatest fixed point where the latter takes its least.
Section~\ref{sec:infinite_branching} describes a condition that is
necessary for the fixed point characterisation of \emph{total}
correctness to be accurate; namely, that there is no infinite
branching. Section~\ref{sec:related} discusses related work, mainly
focussing on how our result extends a similar observation made by
Clarke in 1979~\cite{clarke79}. We begin, in
Section~\ref{sec:preliminaries}, by establishing some preliminary
definitions.

@@

\paragraph{Proof mechanisation} Our main theorem and its accompanying
lemmas have been formalised and proved in the Isabelle theorem prover,
with the exception of Lemma~\ref{lem:longest-trace}, of which we
currently only have a hand proof. The proof relies on K\H{o}nig's
infinity lemma, which we found to be difficult to formalise. The
proof is available online.\footnote{\url{http://www.doc.ic.ac.uk/~jpw48/Partial_Total_Correctness_As_Fixed_Points.thy}}

@@

Clarke~\cite{clarke79} was the first to study the correspondence
between greatest/least fixed points and partial/total correctness. He
remarks, in a 1979 article, that 
%
\begin{quote}the fixedpoints of $\Gamma$ form a complete
lattice under the natural partial ordering on $\pow(\State)$. The top
element of this lattice is the weakest precondition for partial
correctness and the bottom element is the weakest precondition for
total correctness.~\cite[p.~279, footnote]{clarke79} \end{quote}
%
We note three ways in which the current work extends on Clarke's
original observation. First, Clarke works in the setting of a
particular programming language that provides syntax for sequencing,
conditionals, assignment, and recursive calls to parameterless
procedures. In this paper, we need not tackle issues of programming
language syntax, since we work at the level of arbitrary transition
systems. Second, Clarke's work applies only in a deterministic
setting: a program has either a single final state, or none at all. We
allow for highly non-deterministic programs, and only restrict to
finite non-determinism in the case of total correctness. Third,
Clarke's work is in the setting of a big-step semantics; that is, the
meaning of each programming construct maps an initial state directly
to a final state. We work with a small-step semantics (as described by
a transition system), which means that our approach handles parallel
programs without further adaptation. Big-step semantics is well-known
to be unable to handle parallelism, as acknowledged by Clarke, who
remarks that he `[is] currently attempting to extend this fixedpoint
theory to additional programming features such as
parallelism'~\cite[p.~292]{clarke79}. The definition of total
correctness in a big-step and deterministic setting is trivial, and
hence Clarke's $\Gamma$ functional is straightforward -- this perhaps
explains why Clarke consigned the observation we quote above to a mere
footnote. Our functional, on the other hand, which we call $\SafeOne$,
is fairly subtle: experience shows that even minor modifications of
Definition~\ref{defn:safeone} quickly lead to either the least or the
greatest fixed point becoming degenerate.

>>>>

In the context of Hoare Logic~\cite{hoare69}, a (possibly
non-deterministic, and possibly non-terminating) program satisfies a
\emph{total correctness} specification when each of its traces from a
state that satisfies the given precondition reaches a terminal state
satisfying the given postcondition within a finite number of execution
steps. That program satisfies a \emph{partial correctness}
specification when each of its traces either meets the requirement
above or is infinite.

@@

The descriptions above have been deliberately crafted to emphasise a
connection between total/partial correctness and least/greatest fixed
points. This paper seeks to state that connection precisely, and
investigate conditions under which it holds.

Section~\ref{sec:result} provides characterisations of partial and
total correctness, that differ only in that the former takes a
function's greatest fixed point where the latter takes its least.
Section~\ref{sec:infinite_branching} describes a condition that is
necessary for the fixed point characterisation of \emph{total}
correctness to be accurate; namely, that there is no infinite
branching. Section~\ref{sec:related} discusses related work, mainly
focussing on how our result extends a similar observation made by
Edmund Clarke in 1979. We begin, in Section~\ref{sec:preliminaries},
by establishing some preliminary definitions.

@@

\paragraph{Proof mechanisation} Our main theorem and its accompanying
lemmas have been formalised and proved in the Isabelle theorem
prover~\cite{nipkow+02}, with the exception of
Lemma~\ref{lem:longest-trace}, of which we currently only have a hand
proof. The proof relies on K\H{o}nig's infinity lemma, which we found
to be difficult to formalise. The proof is available
online.\footnote{\url{http://www.doc.ic.ac.uk/~jpw48/Partial_Total_Correctness_As_Fixed_Points.thy}}

@@

Edmund Clarke~\cite{clarke79} was probably the first to study the
correspondence between greatest/least fixed points and partial/total
correctness. He remarks, in a 1979 article, that
%
\begin{quote}the fixedpoints of $\Gamma$ form a complete
lattice under the natural partial ordering on $\pow(\State)$. The top
element of this lattice is the weakest precondition for partial
correctness and the bottom element is the weakest precondition for
total correctness.~\cite[p.~279, footnote]{clarke79} \end{quote}
%
We note three ways in which the current work extends on Clarke's
original observation. First, Clarke works in the setting of a
particular programming language that provides syntax for sequencing,
conditionals, assignment, and recursive calls to parameterless
procedures. In this paper, we need not tackle issues of programming
language syntax, since we work at the level of arbitrary transition
systems. Second, Clarke's work applies only in a deterministic
setting: a program has either a single final state, or none at all. We
allow for highly non-deterministic programs, and only restrict to
finite non-determinism in the case of total correctness. Third,
Clarke's work is in the setting of a big-step semantics; that is, the
meaning of each programming construct maps an initial state directly
to a final state. We work with a small-step semantics (as described by
a transition system), which means that our approach handles parallel
programs without further adaptation. Big-step semantics is well-known
to be unable to handle parallelism, as acknowledged by Clarke, who
remarks that he `[is] currently attempting to extend this fixedpoint
theory to additional programming features such as
parallelism'~\cite[p.~292]{clarke79}. The definition of total
correctness in a big-step and deterministic setting is trivial, and
hence Clarke's $\Gamma$ functional is straightforward -- this perhaps
explains why Clarke consigned the observation we quote above to a mere
footnote. Our functional, on the other hand, which we call $\SafeOne$,
is fairly subtle: experience shows that even minor modifications of
Definition~\ref{defn:safeone} quickly lead to either the least or the
greatest fixed point becoming degenerate.
